Deep Gaussian Processes and Hierarchical Modeling with Functional Data: A Case Study in Cosmological Power Spectra

### 2024-08-28 paper/main.tex
update initial details for CAMB/LT runs

In this work, we expand on the previous emulation methods and propose a novel synthesis of methods to quantify uncertainty and predict matter power spectra for different cosmologies using deep Gaussian processes (DGPs). In order to handle multiple realizations of functional output (from the Mira-Titan simulation suite), we use hierarchical and basis functions modeling within our DGP framework to accurately estimate and predict power spectra for different cosmologies. When comparing to Cosmic Emu, we see favorable results for our method. The code to reproduce these results are available at \url{https://github.com/stevewalsh124/dgp.hm}. \textcolor{blue}{Please note: this repository is currently private until we hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

@@

\textcolor{blue}{Annie, please put the newest prior specs here.}

@@

$$
\begin{aligned}
\bar Y|S &\sim \mathcal{N}_n(S,\Sigma_\varepsilon) \\
S &\sim GP\left(\mu, \Sigma_S(x,x')\right)
\end{aligned}
$$

@@

$$
\begin{aligned}
\bar Y|S,W &\sim \mathcal{N}_n(S,\Sigma_\varepsilon) \\
S|W &\sim GP\left(\mu, \Sigma_S(w,w')\right) \\
W &\sim GP\left(0, \Sigma_W(x,x')\right)
\end{aligned}
$$

@@

\textcolor{blue}{Annie: want to add something here? We discuss the elliptical slice sampling of $W$, but otherwise it is plugged in to the methods of the previous section with very few changes.}

@@

We compare our DGP.HM model to competing models with a simulation study, in which case the true curve is known. Following \cite{moran2024dpc}, we consider two functions (1 and 2) each with two variance specifications (settings ``A", ``B", \textcolor{blue}{, we can consider additional homoskedastic cases in the appendix if necessary}). \textcolor{blue}{Put in the details of the two functions, and the two variance specs.} The competing methods we consider are an out-of-the-box DGP fit using the \texttt{deepgp} package \citep{sauer2023active}, a heteroskedastic GP fit using the \texttt{hetGP} package \citep{binois2018practical, binois2021hetgp}, as well as a deep process convolution (DPC) approach, which is utilized within Cosmic Emu \citep{moran2023mira}.

For each simulated function and variance specification, we generate either 5 or 20 functional observations and estimate the underlying true function based on the observations. The metrics we use to evaluate performance are mean squared error (MSE) and the log score, a proper scoring rule which takes both prediction and intervals into consideration \citep{gneiting2007strictly}.

MSE and log score results are shown in Figures \ref{fig:sims_MSE} and \ref{fig:sims_logS} respectively; our DGP.HM model performs favorably against these current benchmark methods.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{sims_logS.png}
    \caption{Boxplot of log scores across two different functions and covariance settings (lower is better). For each case, twenty random batches were simulated.}
    \label{fig:sims_logS}
\end{figure}

@@

Here, we look over the CAMB data to see how our method performs.

>>>>

In this work, we expand on the previous emulation methods and propose a novel synthesis of methods to quantify uncertainty and predict matter power spectra for different cosmologies using deep Gaussian processes (DGPs). In order to handle multiple realizations of functional output (from the Mira-Titan simulation suite), we use hierarchical and basis functions modeling within our DGP framework to accurately estimate and predict power spectra for different cosmologies. When comparing to Cosmic Emu, we see favorable results for our method. The code to reproduce these results are available at \url{https://github.com/stevewalsh124/dgp.hm}. \textcolor{red}{Please note: this repository is currently private until we hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

@@

\textcolor{cyan}{Annie, please put the newest prior specs here.}

@@

\begin{align}
\bar Y|S &\sim \mathcal{N}_n(S,\Sigma_\varepsilon) \\
S &\sim GP\left(\mu, \Sigma_S(x,x')\right)
\end{align}

@@

\begin{align}
\bar Y|S,W &\sim \mathcal{N}_n(S,\Sigma_\varepsilon) \\
S|W &\sim GP\left(\mu, \Sigma_S(w,w')\right) \\
W &\sim GP\left(0, \Sigma_W(x,x')\right)
\end{align}

@@

\textcolor{cyan}{Annie: want to add something here? We discuss the elliptical slice sampling of $W$, but otherwise it is plugged in to the methods of the previous section with very few changes.}

@@

We compare our DGP.HM model to competing models with a simulation study, in which case the true curve is known. Following \cite{moran2024dpc}, we consider two functions ($f_1$ and $f_2$) each with two variance specifications (settings ``4" and ``5", \textcolor{magenta}{we can consider additional homoskedastic cases in the appendix if necessary}).

Here, $f_1$ and $f_2$ are defined as follows:

\begin{align}
  f_1(x) &= m_1 \exp(-u_1x/2) * \cos(wx) - m_1x/5, \quad w=\sqrt{25-[u_1/2]^2} \\
  f_2(x) &= \exp(-m_2[x-3]^2)+\exp(-u_2[x-1]^2)-0.05\sin(8[x-1.9]), \\ 
  \quad x &\in [0,4].
\end{align}

For additional variation, the values of $m_1, m_2, u_1,$ and $u_2$ are random variables where $m_1 \sim Uniform(0.5,1.5), u_1 \sim Uniform(1.5,2.5),$ and $m_2,u_2 \sim Uniform(0.6,1.4)$. For each simulated function and variance specification, we generate 20 values of $m_1, m_2, f_1, f_2$, and conditional on these values, we sample either 5 or 20 functional observations and estimate the underlying true function based on the observations.

The competing methods we consider are an out-of-the-box DGP fit using the \texttt{deepgp} package \citep{sauer2023active}, a heteroskedastic GP fit using the \texttt{hetGP} package \citep{binois2018practical, binois2021hetgp}, as well as a deep process convolution (DPC) approach, which is utilized within Cosmic Emu \citep{moran2023mira}.

The metrics we use to evaluate performance are mean squared error (MSE) and the log score, a proper scoring rule which takes both prediction and intervals into consideration \citep{gneiting2007strictly}. For both metrics, a lower value indicates better performance. MSE and log score results are shown in Figures \ref{fig:sims_MSE} and \ref{fig:sims_logS} respectively; our DGP.HM model performs favorably against these current benchmark methods.

\begin{figure}[t]
    \centering
    \includegraphics[width=6in]{sims_logS.png}
    \caption{Boxplot of log scores across two different functions and covariance settings (lower is better). For each case, twenty random batches were simulated.}
    \label{fig:sims_logS}
\end{figure}

@@

In addition to simulated data, we investigate model fits the for different cosmological datasets (based on linear theory and/or runs from the Code for Anisotropies in the Microwave Background, CAMB). \textcolor{magenta}{We have looked at two sets of model runs, "LT" runs where there is an underlying ``true" value alongside 15 computer model runs, or the "CAMB" runs, where we only have a single output. The performance from the CAMB runs was better (with a 6-dimensional design space), and also did not require obtaining a posterior mean from multiple model runs. When looking at the CAMB runs on the $\Delta$ space, we see it generally is withing $\pm 5\%$ error.}

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_pca_CAMB.png}
    \caption{Proportion of $\Delta_{emu}$ vs. $\Delta_{holdout}$ for model runs 11-42 (excluding 27-28).}   
    \label{fig:pca_camb}
\end{figure}
